---
title: "Introduction to R Workshop"
subtitle: "Getting Started, Data Visualization, and Regression"
author: "Muji Chughtai"
date: today
format:
  beamer:
    theme: default
    aspectratio: 169
    slide-level: 2
    pdf-engine: xelatex
    include-in-header: beamer-header.tex
    keep-tex: false
    incremental: true
---

# Getting Started

## Introduction to R and RStudio

\pause

- R is a free software environment for statistical computing and graphics that can compile and run on a variety of platforms.

  - To download R, go to <https://cran.r-project.org>.

- For a more user-friendly interface, RStudio is a popular integrated development environment (IDE) for R.

- RStudio also allows for powerful document creation using Markdown, which can generate reports, presentations, and dashboards (this presentation was made using a Quarto Markdown in RStudio).

  - To download RStudio, go to <https://www.rstudio.com/products/rstudio/download>.

## Getting Familiar with the RStudio Interface

\pause

There are 4 main panels in the RStudio interface:

\pause

1.  **Source Panel (Top Left)**: This is where you can write and edit your R scripts and markdown files. Toggling between tabs allows you to manage multiple files at once.

2.  **Console Panel (Bottom Left)**: This is where you can directly type and execute R commands. When you run code from the Source panel, the output will also appear here.

3.  **Environment/History Panel (Top Right)**: The Environment tab shows the objects (variables, data frames, functions, etc.) currently in your R session.

4.  **Files/Plots/Packages/Help/Viewer Panel (Bottom Right)**: This panel has multiple tabs for managing files, viewing plots, installing and loading packages, accessing R help documentation, and displaying web content.

## Using Markdown

\pause

- R Markdown is a simple formatting syntax for authoring documents. It allows you to combine text, code, and output in a single document. You can create reports, presentations, and dashboards using R Markdown.

- In R Markdown, you write code in **chunks** that are enclosed by triple backticks and curly braces.

  - You can create R chunks using the green "c" button in the RStudio toolbar or by typing in three backticks followed by `{r}`.

  - If you want to write in other languages in the code chunk, you can specify the language in the curly braces (e.g., `{python}`, `{bash}`, etc.).

- I suggest you do your analyses in R Markdown documents, as it allows you to keep your code and results together in a reproducible format that can be easily exported to various formats (HTML, PDF, Word, etc.)!

## Staying Organized

\pause

- One of the biggest issues that people face when programming (in R) is keeping track of code, files, and output.

- To help with this, I encourage you to create a dedicate `RProject` for each project you work on. This will keep all of your files in one folder, and within this folder you can have subfolders for `Data`, `Analyses`, `Figures`, etc.

- The `here` package is a great tool to help you manage file paths in your R projects. It allows you to use relative paths instead of absolute paths, which makes your code more portable and easier to share.

# Data Visualization

## Creating a Directory

\pause

Let's start programming in R by creating a small project where we will visualize and analyze some data from using data from the World Bank in 2016.

\pause

- Open RStudio and create a new project (and corresponding folder) by navigating from the menu to `File > New Project > New Directory > New Project`, and name your project whatever you like and save it in a location you prefer.

- Within that project folder, create a new R Markdown document by navigating to `File > New File > R Markdown`, and you can do this from either the menu or the bottom right panel. Name it whatever you like!

## Installing and Loading Packages

\pause

- Let's load in two packages that we will use for this workshop: `tidyverse` and `here`.

  - `Tidyverse` is a collection of R packages that are designed for data science. It includes packages for data manipulation, visualization, and modeling.

  - `Here` is a package that helps you manage file paths in your R projects.

- To load in packages in R you use the `library()` function:

\pause

```{r, echo = TRUE}
library(tidyverse)
library(here)
```

## Installing and Loading Packages

- This code wont work if you don't have the packages installed. To install them, you can use the `install.packages()` function, and you should put this code chunk above the `library()` code chunk in your R Markdown document:

\pause

```{r, echo = TRUE}
#| eval: false
install.packages("tidyverse", repos = "https://cloud.r-project.org/")
install.packages("here", repos = "https://cloud.r-project.org/")
```

\pause

- Newer versions of RStudio will also allow you to download packages from the top menu if you don't have them.

- You can also directly install packages from the console by typing in the same code above (when you run a chunk it will run in the console).

## Installing and Loading Packages

\pause

- You can make comments in your code using the `#` symbol within a chunk. Comments are ignored by R when running code, and they are a great way to explain what your code is doing:

\pause

```{r, echo = TRUE}
# loading in required packages
library(tidyverse)
library(here)
```

\pause

- If you use the `#` symbol outside of a chunk, it will be treated as a header in your R Markdown document. The more `#` symbols you use, the smaller the header will be.

- These headings will transfer into the output document that you render for easy navigation in the html, pdf, or word document.

## Loading in the Data

\pause

- Let's load in the data from this project, which is a csv file that contains data from the World Bank in 2016.

- The data and codebook is available on my github at <https://github.com/mujtabachughtai/2026-R-Workshop>.

- Download the data (called `WB.2016.csv`), create a `Data` subfolder in this project's directory, and save the data in that folder.

## Loading in the Data

- Then you should be able to load in the data using the `here` package, where you can name your new dataframe `WorldBank`

\pause

```{r, echo = TRUE}
WorldBank <- read.csv(
  here("Data", "WB.2016.csv"),
  fileEncoding = "Windows-1252"
)
```

\pause

- Note: You normally don't need the `fileEncoding` argument, but this dataset has some special characters that require it.

- You should now be able to view your data in the `Environment` tab in the top right panel of RStudio.

## Getting a Feel for the Data

\pause

- When you look at the dataframe in RStudio, you can see what variables we have to explore.

- For this workshop, we will be focusing on the following variables currently in the `WorldBank` dataframe:

  - `Country`: The name of the country
  - `Code`: A 3-letter code for each country
  - `Population`: The population of the country
  - `GNI`: Gross National Income per capita

- And we're going to make some variables of our own, too.

## Getting a Feel for the Data

- Let's take a look at the first few rows of these key variables. We will use the `knitr::kable` and `head` functions to do so.

```{r, echo = TRUE}
#| eval: false
knitr::kable(head(
  WorldBank[, c("Country", "Code", "Population", "GNI")]),
  caption = "First 5 Rows of our Key Variables from the
2016 World Bank Data")
```

## Getting a Feel for the Data

```{r}
#| eval: true
knitr::kable(head(
  WorldBank[, c("Country", "Code", "Population", "GNI")]),
  caption = "First 5 Rows of our Key Variables from the 2016 World Bank Data")
```

## Getting a Feel for the Data

\pause

- It looks like there is some misingness in data, which is something we should keep in mind as we are doing our analyses.

- We can check the means of our key variables using the R's build in mean function:

\pause

```{r, echo = TRUE}
# mean of Population
mean(WorldBank$Population, na.rm = TRUE)
```

\pause

```{r, echo = TRUE}
# mean of GNI
mean(WorldBank$GNI, na.rm = TRUE)
```

## Getting a Feel for the Data

\pause

- We can also look at the distributions of these variables using `geom_density` from `ggplot`:

\pause

```{r, echo = TRUE}
#| eval: false
WorldBank %>%
  ggplot(aes(x = Population)) +
  geom_density()
```

\pause

- This can equivalently be written as

\pause

```{r, echo = TRUE}
#| eval: false
ggplot(aes(x = Population), data = WorldBank) +
  geom_density()
```

## Getting a Feel for the Data

- `ggplot` objects add layers to the plot using the `+` operator, and you can add as many layers as you want.

- Let's make this plot cleaner by changing the theme and additing title and axis labels:

\pause

```{r, echo = TRUE}
#| eval: false
WorldBank %>%
  ggplot(aes(x = Population)) +
  geom_density() +
  theme_minimal() +
  labs(
    title = "Distribution of Population from 2016 World Bank Data",
    x = "Population",
    y = "Density"
  )
```

## Getting a Feel for the Data

```{r, fig.width = 12, fig.height = 6.75, out.width = "95%", fig.align = "center"}
#| eval: true
WorldBank %>%
  ggplot(aes(x = Population)) +
  geom_density() +
  theme_minimal(base_size = 22) +
  labs(
    title = "Distribution of Population from 2016 World Bank Data",
    x = "Population",
    y = "Density"
  )
```

## Making Transformations

- The plot looks great, but the data is very heavily right skewed:

  - This is because the vast majority of countries that have small populations, while there are a few that have very large ones.

- We can log-transform the data to make it more normally distributed, which is a common technique in statistics. This will also help us with our regression analyses later on.

## Making Transformations

\pause

- The `mutate` function allows you to create new variables based on the data from other variables. Let's use this function to create a new variable called `logPopulation` that is just the natural log of the `Population` variables, and we will recreate the `WorldBank` dataframe with this new variable:

\pause

```{r, echo = TRUE}
WorldBank <- WorldBank %>%
  mutate(logPopulation = log(Population))
```

## Making Transformations

\pause

- Now we can recreate the density plot using `logPopulation`:

\pause

```{r, echo = TRUE}
#| eval: false
WorldBank %>%
  ggplot(aes(x = logPopulation)) +
  geom_density() +
  theme_minimal() +
  labs(
    title = "Distribution of
log-transfored Population from 2016 World Bank Data",
    x = "log(Population)",
    y = "Density"
  )
```

## Making Transformations

```{r, fig.width = 12, fig.height = 6.75, out.width = "95%", fig.align = "center"}
#| eval: true
WorldBank %>%
  ggplot(aes(x = logPopulation)) +
  geom_density() +
  theme_minimal(base_size = 22) +
  labs(
    title = "Distribution of log-transformed Population from 2016 World Bank Data",
    x = "log(Population)",
    y = "Density"
  )
```

## Making Transformations

\pause

- Let's do the same for GNI

\pause

```{r, echo = TRUE}
#| eval: false
WorldBank <- WorldBank %>%
  mutate(logGNI= log(GNI))

WorldBank %>%
  ggplot(aes(x = logGNI)) +
  geom_density() +
  theme_minimal(base_size = 22) +
  labs(
    title = "Distribution of log-transformed
GNI from 2016 World Bank Data",
    x = "log(GNI)",
    y = "Density"
  )
```

## Making Transformations

```{r, fig.width = 12, fig.height = 6.75, out.width = "95%", fig.align = "center"}
#| eval: true
WorldBank <- WorldBank %>%
  mutate(logGNI= log(GNI))

WorldBank %>%
  ggplot(aes(x = logGNI)) +
  geom_density() +
  theme_minimal(base_size = 22) +
  labs(
    title = "Distribution of log-transformed GNI from 2016 World Bank Data",
    x = "GNI",
    y = "Density"
  )
```

## Plotting Associations

\pause

- Now we can compare the relationship between a country's GNI and its population by plotting `logPopulation` versus `logGNI`:

- We do this by using `geom_point()`.

\pause

```{r, echo = TRUE}
#| eval: false

WorldBank %>%
  ggplot(aes(x = logGNI, y = logPopulation)) +
  geom_point(alpha = 0.5) +
  theme_minimal(base_size = 22) +
  labs(
    title = "Country GNI vs Population from 2016 World Bank Data",
    x = "log(Population)",
    y = "log(GNI)"
  )
```

## Plotting Associations

```{r, fig.width = 12, fig.height = 6.75, out.width = "95%", fig.align = "center"}
#| eval: true

WorldBank %>%
  ggplot(aes(x = logGNI, y = logPopulation)) +
  geom_point(alpha = 0.5) +
  theme_minimal(base_size = 22) +
  labs(
    title = "Country GNI vs Population from 2016 World Bank Data",
    x = "log(Population)",
    y = "log(GNI)"
  )
```

## Plotting Associations

\pause

- We can also add a line of best fit to this plot to see if this relationship tends to be negative or positive (and we will formally test this relationship later using linear regression)

\pause

```{r, echo = TRUE}
#| eval: false

WorldBank %>%
  ggplot(aes(x = logGNI, y = logPopulation)) +
  geom_point(alpha = 0.5) +
  geom_smooth(method = "lm", se = FALSE, color = "lightblue") +
  theme_minimal(base_size = 22) +
  labs(
    title = "Country GNI vs Population from 2016 World Bank Data",
    x = "log(Population)",
    y = "log(GNI)"
  )
```

## Plotting Associations

```{r, fig.width = 12, fig.height = 6.75, out.width = "95%", fig.align = "center"}
#| eval: true

WorldBank %>%
  ggplot(aes(x = logGNI, y = logPopulation)) +
  geom_point(alpha = 0.5) +
  geom_smooth(method = "lm", se = FALSE, color = "lightblue") +
  theme_minimal(base_size = 22) +
  labs(
    title = "Country GNI vs Population from 2016 World Bank Data",
    x = "log(Population)",
    y = "log(GNI)"
  )
```

## Plotting Associations

\pause

- It looks like there is a slight negative relationship between a country's GNI and its population, but before we formally test this relationship, we can try to incorporate more information about the underlying country datapoints into the plot.

- Specifically, we can color countries based on the continent that they come from to get an initial sense on if the relationship between a country's population and it GNI might differ across different continents.

- However, when looking at the current `WorldBank` dataframe, there is no continent variable.

- We could use `mutate` again to create this new `continent` variable, but there are too many countries to assess the value of to create the corresponding continent.

## Combining Dataframes

\pause

- Instead, we could take another dataframe (which I have created and is available on my github) that has the continents for each country, and we could combine it to our current `WorldBank` dataframe.

- To combine dataframes, you have to have a variable that is shared in both so that rows are combined when the values in that column are shared.

- For us, that would be the `Country` variable

- Let's load in the `CountryContinents.csv` file into our workspace and combine it into a new `WorldBank` dataframe using `left_join`

## Combining Dataframes

```{r, echo = TRUE}
CountryContinents <- read.csv(
  here("Data", "CountryContinents.csv"),
  fileEncoding = "Windows-1252"
)

WorldBank <- WorldBank %>%
  left_join(CountryContinents, by = "Country")
```

      
## Computing Summary Statistics

- When you click on the dataframe, you should now see the `Continent` column.

- With this new grouping variable, we can also calculate summary statistics for each group separately. Let's do this before remaking the `logPopulation` vs `logGNI` plot with `Continent`.

- To calculate summary statistics for each group, we can use the `group_by` and `summarise`:

## Computing Summary Statistics

```{r, echo = TRUE}
#| eval: false
WorldBank %>%
  group_by(Continent) %>%
  summarise(
    n = n(),
    mean_logGNI = mean(logGNI, na.rm = TRUE),
    mean_logPopulation = mean(logPopulation, na.rm = TRUE)
  ) %>%
  knitr::kable(caption = "Summary Statistics by Continent",
               col.names = c("Continent", "Number of Countries",
                             "Mean log(GNI)", "Mean log(Population)"))
```

## Computing Summary Statistics

```{r}
#| eval: true
WorldBank %>%
  filter(!is.na(Continent)) %>%
  group_by(Continent) %>%
  summarise(
    n = n(),
    mean_logGNI = mean(logGNI, na.rm = TRUE),
    mean_logPopulation = mean(logPopulation, na.rm = TRUE)
  ) %>%
  knitr::kable(caption = "Summary Statistics by Continent",
               col.names = c("Continent", "Number of Countries",
                             "Mean log(GNI)", "Mean log(Population)"))
```

## Plotting 3 Variables

\pause

- Looks great! Now we can remake the `logPopulation` vs `logGNI` plot and color the points (and lines) by `Continent`:

\pause

```{r, echo = TRUE}
#| eval: false
WorldBank %>%
  filter(!is.na(Continent)) %>%
  ggplot(aes(x = logPopulation, y = logGNI, color = Continent)) +
  geom_point(alpha = 0.5) +
  theme_minimal(base_size = 22) +
  labs(
    title = "Country GNI vs Population from 2016 World Bank Data",
    x = "log(Population)",
    y = "log(GNI)"
  )
```

## Plotting 3 Variables

```{r, fig.width = 12, fig.height = 6.75, out.width = "95%", fig.align = "center"}
#| eval: true
WorldBank %>%
  filter(!is.na(Continent)) %>%
  ggplot(aes(x = logPopulation, y = logGNI, color = Continent)) +
  geom_point(alpha = 0.5) +
  geom_smooth(method = "lm", se = FALSE) +
  theme_minimal(base_size = 22) +
  labs(
    title = "Country GNI vs Population from 2016 World Bank Data",
    x = "log(Population)",
    y = "log(GNI)"
  )
```

## Plotting 3 Variables

\pause

- It looks like the relationship between `logPopulation` and `logGNI` might differ across continents! To make this even clearer, we can make 5 separate mini plots for this relationship, one for each continent:

\pause

```{r, echo = TRUE}
#| eval: false
WorldBank %>%
  filter(!is.na(Continent)) %>%
  ggplot(aes(x = logPopulation, y = logGNI)) +
  facet_wrap(~Continent) +
  geom_point(alpha = 0.5) +
  geom_smooth(method = "lm", se = FALSE) +
  theme_minimal(base_size = 22) +
  labs(
    title = "Country GNI vs Population from 2016 World Bank Data",
    x = "log(Population)",
    y = "log(GNI)"
  )
```

## Plotting 3 Variables

```{r, fig.width = 12, fig.height = 6.75, out.width = "95%", fig.align = "center"}
#| eval: true
WorldBank %>%
  filter(!is.na(Continent)) %>%
  ggplot(aes(x = logPopulation, y = logGNI)) +
  facet_wrap(~Continent) +
  geom_point(alpha = 0.5) +
  geom_smooth(method = "lm", se = FALSE) +
  theme_minimal(base_size = 22) +
  labs(
    title = "Country GNI vs Population from 2016 World Bank Data",
    x = "log(Population)",
    y = "log(GNI)"
  )
```

## Saving Plots

\pause

- This is a nice plot! We can save it to our `Figures` folder in our directory (make sure you have a folder named `Figures` in your working directory) by saving the `ggplot` object and using `ggsave`:

\pause

```{r, echo = TRUE}
GNI_Population_by_Continent <- WorldBank %>%
  filter(!is.na(Continent)) %>%
  ggplot(aes(x = logPopulation, y = logGNI)) +
  facet_wrap(~Continent) +
  geom_point(alpha = 0.5) +
  geom_smooth(method = "lm", se = FALSE) +
  theme_minimal(base_size = 22) +
  labs(
    title = "Country GNI vs Population from 2016 World Bank Data",
    x = "log(Population)",
    y = "log(GNI)"
  )
```

## Saving Plots

```{r, echo = TRUE}
ggsave(
  GNI_Population_by_Continent,
  filename = here("Figures", "GNI_Population_by_Continent.png"),
  width = 32,
  height = 24,
  units = "in",
  dpi = 300
)
```

# Linear Regression

## Theory

```{=latex}
\begin{definitionblock}
Ordinary least squares regression estimates the intercept and slope by minimizing the sum of the squared residuals:

\[
(\hat{\beta}_0,\hat{\beta}_1)
=
\underset{\beta_0,\beta_1}{\operatorname{arg\,min}}
\sum_{i=1}^{n}
\left[y_i-(\beta_0+\beta_1x_i)\right]^2.
\]

The resulting regression equation is

\[
\hat{y}_i=\hat{\beta}_0+\hat{\beta}_1x_i.
\]
\end{definitionblock}
```

## Implementation

\pause

- You can fit linear models very easily in `R` using the `lm()` function

- Given that it seems like the relationship between a country's GNI and its population might differ across continents, we will fit an interaction model.

  - This will also take into account the likely main effect of `Continent` on our other two variables.
  
## Implementation

\pause

```{r, echo = TRUE}
#| eval: false
FullModel <- lm(logGNI ~ logPopulation * Continent, data = WorldBank)
summary(FullModel)
```

\pause

- And we can further examine interactions by using the `simple_slopes` function:

\pause

```{r, echo = TRUE}
#| eval: false
library(reghelper)
simple_slopes(FullModel, pred = "logPopulation", modx = "Continent")
```

## Interpretation

- From this output, we can make a few interpretations

  - Even after controlling for the main effect of `Continent`, there is still a significant negative relationship between a country's population and its GNI.
  
  - When Africa is the reference group, the slope of the relationship between a country's population and its GNI is -0.19
  
  - No interactions are significant: the slope does not significantly change from this reference group for any of the other continents.
  
  - However, when examining the slopes for different continents, the association between a country's population and its GNI is only significant for Africa.
  

# One Last Fun Plot

- You can add a lot of different variables to `ggplot` object, but at risk of making the plots hard to interpret (there are only so many ways to manipulate a visual object!).

- Let's try to add a bit for information to our plot by changing the points to the 3-character country code and by changing the color of the points based on how much fossil fuels are used in the country (as a percentage of total energy use).

## Plotting 4 Variables

```{r, echo = TRUE}
#| eval: false

WorldBank %>%
  filter(!is.na(Continent)) %>%
  ggplot(aes(x = logPopulation, y = logGNI, color = FossilPct)) +
  geom_text(aes(label = Code), size = 2) +
  facet_wrap(~Continent) +
  geom_smooth(method = "lm", se = FALSE) +
  theme_minimal(base_size = 14) +
  labs(
    title = "Country GNI vs Population from 2016 World Bank Data",
    x = "log(Population)",
    y = "log(GNI)",
    color = "% of Energy from Fossil Fuels"
  )
```

## Plotting 4 Variables

```{r, fig.width = 12, fig.height = 6.75, out.width = "95%", fig.align = "center"}
#| eval: true

WorldBank %>%
  filter(!is.na(Continent)) %>%
  ggplot(aes(x = logPopulation, y = logGNI, color = FossilPct)) +
  geom_text(aes(label = Code), size = 2) +
  facet_wrap(~Continent) +
  geom_smooth(method = "lm", se = FALSE) +
  theme_minimal(base_size = 14) +
  labs(
    title = "Country GNI vs Population from 2016 World Bank Data",
    x = "log(Population)",
    y = "log(GNI)",
    color = "% of Energy from Fossil Fuels"
  )
```

## Plotting 4 Variables

\pause

- And we can save as normal:

\pause

```{r, echo = TRUE}
Pop_GNI_Fossil <- WorldBank %>%
  filter(!is.na(Continent)) %>%
  ggplot(aes(x = logPopulation, y = logGNI, color = FossilPct)) +
  geom_text(aes(label = Code), size = 2) +
  facet_wrap(~Continent) +
  geom_smooth(method = "lm", se = FALSE) +
  theme_minimal(base_size = 14) +
  labs(
    title = "Country GNI vs Population from 2016 World Bank Data",
    x = "log(Population)",
    y = "log(GNI)",
    color = "% of Energy from Fossil Fuels"
  )
```

## Plotting 4 Variables

```{r, echo = TRUE}
ggsave(
  Pop_GNI_Fossil,
  filename = here("Figures", "Pop_GNI_Fossil.png"),
  width = 32,
  height = 24,
  units = "in",
  dpi = 300
)
```

# Summary of Exercises

\pause

- In this workshop, we learned how to:

  - Get started with R and RStudio
  - Use R Markdown to create reproducible documents
  - Load in data and explore it using summary statistics and visualizations
  - Create new variables and transform existing ones
  - Combine dataframes
  - Fit linear regression models and interpret the results
  - Save plots for future use

