---
title: "Introduction to R Workshop"
subtitle: "Getting Started, Data Visualization, and Regression"
author: "Muji Chughtai"
date: today
format:
  beamer:
    theme: default
    aspectratio: 169
    slide-level: 2
    pdf-engine: xelatex
    include-in-header: beamer-header.tex
    keep-tex: false
    incremental: true
---

# Getting Started

## Introduction to R and RStudio

\pause

- R is a free software environment for statistical computing and graphics that can compile and run on a variety of platforms.

  - To download R, go to <https://cran.r-project.org>.

- For a more user-friendly interface, RStudio is a popular integrated development environment (IDE) for R.

- RStudio also allows for powerful document creation using Markdown, which can generate reports, presentations, and dashboards (this presentation was made using a Quarto Markdown in RStudio).

  - To download RStudio, go to <https://www.rstudio.com/products/rstudio/download>.

## Getting Familiar with the RStudio Interface

\pause

There are 4 main panels in the RStudio interface:

\pause

1.  **Source Panel (Top Left)**: This is where you can write and edit your R scripts and markdown files. Toggling between tabs allows you to manage multiple files at once.

2.  **Console Panel (Bottom Left)**: This is where you can directly type and execute R commands. When you run code from the Source panel, the output will also appear here.

3.  **Environment/History Panel (Top Right)**: The Environment tab shows the objects (variables, data frames, functions, etc.) currently in your R session.

4.  **Files/Plots/Packages/Help/Viewer Panel (Bottom Right)**: This panel has multiple tabs for managing files, viewing plots, installing and loading packages, accessing R help documentation, and displaying web content.

## Using Markdown

\pause

- R Markdown is a simple formatting syntax for authoring documents. It allows you to combine text, code, and output in a single document. You can create reports, presentations, and dashboards using R Markdown.

- In R Markdown, you write code in **chunks** that are enclosed by triple backticks and curly braces.

  - You can create R chunks using the green "c" button in the RStudio toolbar or by typing in three backticks followed by `{r}`.
  
  - If you want to write in other languages in the code chunk, you can specify the language in the curly braces (e.g., `{python}`, `{bash}`, etc.).
  
- I suggest you do your analyses in R Markdown documents, as it allows you to keep your code and results together in a reproducible format that can be easily exported to various formats (HTML, PDF, Word, etc.)!

## Staying Organized

\pause

- One of the biggest issues that people face when programming (in R) is keeping track of code, files, and output.

- To help with this, I encourage you to create a dedicate `RProject` for each project you work on. This will keep all of your files in one folder, and within this folder you can have subfolders for `Data`, `Analyses`, `Figures`, etc.

- The `here` package is a great tool to help you manage file paths in your R projects. It allows you to use relative paths instead of absolute paths, which makes your code more portable and easier to share.

# Data Visualization

## Creating a Directory

\pause

Let's start programming in R by creating a small project where we will visualize and analyze some data from using data from the World Bank in 2016.

\pause

- Open RStudio and create a new project (and corresponding folder) by navigating from the menu to `File > New Project > New Directory > New Project`, and name your project whatever you like and save it in a location you prefer.

- Within that project folder, create a new R Markdown document by navigating to `File > New File > R Markdown`, and you can do this from either the menu or the bottom right panel. Name it whatever you like!

## Installing and Loading Packages

\pause

- Let's load in two packages that we will use for this workshop: `tidyverse` and `here`.

  - `Tidyverse` is a collection of R packages that are designed for data science. It includes packages for data manipulation, visualization, and modeling.
  
  - `Here` is a package that helps you manage file paths in your R projects.
  
- To load in packages in R you use the `library()` function:

\pause

```{r, echo = TRUE}
library(tidyverse)
library(here)
```

## Installing and Loading Packages

- This code wont work if you don't have the packages installed. To install them, you can use the `install.packages()` function, and you should put this code chunk above the `library()` code chunk in your R Markdown document:

\pause

```{r, echo = TRUE}
#| eval: false
install.packages("tidyverse", repos = "https://cloud.r-project.org/")
install.packages("here", repos = "https://cloud.r-project.org/")
```

\pause

- Newer versions of RStudio will also allow you to download packages from the top menu if you don't have them.

- You can also directly install packages from the console by typing in the same code above (when you run a chunk it will run in the console).


## Installing and Loading Packages

\pause

- You can make comments in your code using the `#` symbol within a chunk. Comments are ignored by R when running code, and they are a great way to explain what your code is doing:

\pause

```{r, echo = TRUE}
# loading in required packages
library(tidyverse)
library(here)
```

\pause

- If you use the `#` symbol outside of a chunk, it will be treated as a header in your R Markdown document. The more `#` symbols you use, the smaller the header will be.

- These headings will transfer into the output document that you render for easy navigation in the html, pdf, or word document.

## Loading in the Data

\pause

- Let's load in the data from this project, which is a csv file that contains data from the World Bank in 2016.

- The data and codebook is available on my github at <https://github.com/mujtabachughtai/2026-R-Workshop>.

- Download the data (called `WB.2016.csv`), create a `Data` subfolder in this project's directory, and save the data in that folder.

## Loading in the Data

- Then you should be able to load in the data using the `here` package, where you can name your new dataframe `WorldBank`

\pause

```{r, echo = TRUE}
WorldBank <- read.csv(
  here("Data", "WB.2016.csv"),
  fileEncoding = "Windows-1252"
)
```

\pause

- Note: You normally don't need the `fileEncoding` argument, but this dataset has some special characters that require it.

- You should now be able to view your data in the `Environment` tab in the top right panel of RStudio.

## Getting a Feel for the Data

\pause

- When you look at the dataframe in RStudio, you can see what variables we have to explore.

- For this workshop, we will be focusing on the following variables currently in the `WorldBank` dataframe:

  - `Country`: The name of the country
  - `Code`: A 3-letter code for each country
  - `Population`: The population of the country
  - `GNI`: Gross National Income per capita
  
- And we're going to make some variables of our own, too.

## Getting a Feel for the Data
- Let's take a look at the first few rows of these key variables. We will use the `knitr::kable` and `head` functions to do so.

```{r, echo = TRUE}
knitr::kable(head(
  WorldBank[, c("Country", "Code", "Population", "GNI")]))
```

## Getting a Feel for the Data

\pause

- It looks like there is some misingness in data, which is something we should keep in mind as we are doing our analyses.

- We can check the means of our key variables using the R's build in mean function:

\pause

```{r, echo = TRUE}
# mean of Population
mean(WorldBank$Population, na.rm = TRUE)
```

\pause

```{r, echo = TRUE}
# mean of GNI
mean(WorldBank$GNI, na.rm = TRUE)
```

## Getting a Feel for the Data

\pause

- We can also look at the distributions of these variables using `geom_density` from `ggplot`:

\pause

```{r, echo = TRUE}
#| eval: false
WorldBank %>%
  ggplot(aes(x = Population)) +
  geom_density()
```

\pause

- This can equivalently be written as

\pause


```{r, echo = TRUE}
#| eval: false
ggplot(aes(x = Population), data = WorldBank) +
  geom_density()
```


## Getting a Feel for the Data

- `ggplot` objects add layers to the plot using the `+` operator, and you can add as many layers as you want.

- Let's make this plot cleaner by changing the theme and additing title and axis labels:

\pause

```{r, echo = TRUE}
#| eval: false
WorldBank %>%
  ggplot(aes(x = Population)) +
  geom_density() +
  theme_minimal() +
  labs(
    title = "Distribution of Population from 2016 World Bank Data",
    x = "Population",
    y = "Density"
  )
```

## Getting a Feel for the Data

```{r, fig.width = 24, fig.height = 16, fig.align = "center"}
#| eval: true
WorldBank %>%
  ggplot(aes(x = Population)) +
  geom_density() +
  theme_minimal(base_size = 36) +
  labs(
    title = "Distribution of Population from 2016 World Bank Data",
    x = "Population",
    y = "Density"
  )
```

## Getting a Feel for the Data

- The plot looks great, but the data is very heavily right skewed:

  - This is because the vast majority of countries that have small populations, while there are a few that have very large ones.

- We can log-transform the data to make it more normally distributed, which is a common technique in statistics. This will also help us with our regression analyses later on.


## Theorem

For all real numbers $x$ and $y$, we have $(x+y)^2 = x^2 + 2xy + y^2$
:::

## Academic Environments

Use `::: {.block}`, `::: {.alertblock}`, `::: {.exampleblock}` for Pandoc-native Beamer blocks. Use raw LaTeX (```` ```{=latex} ````) for the custom orange/gray environments.

```{=latex}
\begin{definitionblock}
A \textbf{group} $(G, \cdot)$ is a set with a binary operation satisfying
closure, associativity, identity, and invertibility.
\end{definitionblock}

\begin{proofblock}
By hypothesis every element of $H \leq G$ satisfies $g^n = e$ for some $n$.
Therefore $H$ is generated by $g^k$ for the smallest such $k$.
\end{proofblock}
```

::: alertblock
## Important

This theorem requires the axiom of choice for infinite collections.
:::

## More Environments

::: exampleblock
## Example

The integers $(\mathbb{Z}, +)$ form an abelian group under addition.
:::

```{=latex}
\begin{lemmablock}
Every subgroup of a cyclic group is cyclic.
\end{lemmablock}

\begin{corollaryblock}
If $|G| = p$ is prime then $G \cong \mathbb{Z}/p\mathbb{Z}$.
\end{corollaryblock}
```

## Lists and Structure

- First point with important information
- Second point to demonstrate spacing
- Third point showing the style

1.  Numbered lists also work well
2.  With proper spacing
3.  And clean formatting

## Two Column Layout

::::: columns
::: {.column width="50%"}
**Left Column**

- Item one
- Item two
- Item three
:::

::: {.column width="50%"}
**Right Column**

Content on the right side with matching styling
:::
:::::

## Code Example

```{r, echo = TRUE}
# R code with syntax highlighting
library(ggplot2)

ggplot(mtcars, aes(x = wt, y = mpg)) +
  geom_point() +
  theme_minimal()
```
